### 2015-12-08 paper/1-introduction.tex
First rewrite of introduction

%? use paragraphs for layout instead of (dense) flowing text ?
%TODO SIZE: approx. 1-2 column, every TODO is approx. one paragraph
\subsection{DEVS}
%TODO give introduction to the field: DEVS is used for discrete event simulation and is a common basis for most other formalisms
The family of DEVS \cite{DEVSbase} formalisms serve as a common basis for most other discrete event formalisms. Of interest in this paper are the 2 key formalisms: Dynamic Structured \cite{DSDEVS} and Parallel \cite{ParallelDEVS} and their implementation. This project uses the Direct Connection \cite{SymbolicFlattening} algorithm, so from a simulation kernel's perspective only Atomic Models exist linked to each other by ports.

\subsection{Parallel computing}
%TODO parallel computing is necessary with more and more complex systems, blablabla (usual explanation)
Parallel execution of a PDEVS simulation can lower overall runtime and increase the bound on the state space, thereby enabling simulation of more complex systems in the same time-frame.
While the shared memory parallelism offered by most modern hardware does not raise the state space bounds, it can reduce the runtime and offers more direct communication and control between entities involved in synchronization compared to distributed simulation. \\

\subsection{Motivation}
%TODO give motivation (adevs is good, but only conservative; exploit new parallelism features of C++11 too)
Adevs \cite{adevs} offers a very fast conservative synchronized shared memory PDEVS simulator, but no optimistic synchronized variant. 
The latter can be significantly faster, especially in simulations where the runtime behaviour of the simulation is hard to predict. \\
The matured parallelism features of C++11 were used in this project with the dual aim of writing standard-compliant (and thus portable) code and without losing access to powerful low-level threading primitives. % may need another pass on this one :(

\subsection{Solution} 
%TODO give our solution, motivating the need for a different kernel, instead of modifying another kernel (like adevs or PythonPDEVS)
The usage of the Direct connection algorithm makes reusing adevs kernel hard. Devs-ex-machina (dxex) is based on PythonPDEVS \cite{PythonPDEVS}(PyPDEVS) but differs in design and implementation.
The core aim of the project is to offer a deterministic simulation kernel where the simulation modeller is shielded as much as possible from the kernel implementation, without sacrificing performance. As in PyPDEVS, a model need be written only once for use in the different simulation kernels (with the exception of a non-trivial lookahead).\\
The tracing framework from PyPDEVS was ported to allow optional verification of simulations. 
\subsection{Time}
The DEVS formalisms have $\mathbb{R}$ as time base, but any implementation has to decide on an enumerable representation of time. In principle any type with well defined operators can be used as template parameter, but from a performance point of view a type fitting in a machine word offers obvious advantages. In practice, dxex defaults to word sized unsigned integer.
Although this significantly reduces the possible range of virtual time, it avoids approximation errors. The notion of $\epsilon$ is trivial for integer, unlike floating point.\\
Our time representation is extended with a causality field with a range equal or greater than the maximum nr of models in a single kernel ($2^{48}$ by default). If A and B are imminent at time t then $t[1]_a < t[1]_b \oplus t[1]_b < t[1]_a$, while $t[0]_a == t[0]_b == t[0]$. This allows a deterministic ordering of concurrent events. % execution not so much as tracing

>>>>

%TODO restructure and let it "flow" better
\textsf{DEVS}~\cite{CDEVS} is a popular formalism for modelling complex dynamic systems using a discrete-event abstraction.
In fact, it can serve as a simulation ``assembly language'' to which models in other formalisms can be mapped~\cite{DEVSbase}.
A number of tools have been constructed by academia and industry that allow the modelling and simulation of \textsf{DEVS} models.

With the ever increasing complexity of simulation models, parallel simulation becomes necessary to perform the simulation within reasonable time bounds.
And while \textsf{Parallel DEVS}~\cite{PDEVS} was introduced to increase parallelism, this is often still insufficient.
Several synchronization algorithms from the discrete event simulation community~\cite{FujimotoBook} have been applied to \textsf{DEVS} simulation in particular.
While several parallel \textsf{DEVS} simulation kernels exist, they often restrict themselves to a single synchronization protocol.
This is a logical choice, as these synchronization protocols have nearly no commonalities.
But the exact reason for different synchronization protocols, is that their distinct nature makes them very applicable to specific situations.
As such, current \textsf{DEVS} simulation kernels always allow parallel simulation, but good performance can only be expected from some models.

Users that wish to simulate a set of distinct models, with different synchronization protocols, are therefore out of luck:
Either they accept the lower performance for some of the models, or they use two distinct simulation kernels.
Neither of these options is acceptable for the simulation of complex models, as the performance can become arbitrary worse, and the simulation kernels can diverge significantly.

In this paper, we introduce DEVS-Ex-Machina (``dxex''), our simulation tool which offers multiple synchronization protocols: no synchronization (sequential execution), conservative synchronization, or optimistic synchronization.
The selected synchronization protocol is transparant to the simulated model.
Users should merely determine, at the start of simulation, which protocol they wish to use.
Our tool is implemented in C++11, to increase both performance and portability across different platforms.

A simple model is used to prove that the ideal synchronization algorithm is indeed dependent on model details.
In our case, a single parameter of the model can have a significant impact on simulation performance, and especially on which is the ideal synchronization protocol.
In order to show that this flexibility does not counter our performance, we compare our tool to Adevs, currently one of the fastest \textsf{DEVS} simulation tools available~\cite{PythonPDEVS1}.

We introduce the necessary background for this paper in Section~\ref{sec:2-background}.
Section~\ref{sec:3-features} elaborates on our features and design.
In Section~\ref{sec:4-performance}, we evaluate performance of our tool by comparing its different synchronization protocols.
Related work is discussed in Section~\ref{sec:5-related-work}.
Section~\ref{sec:6-conclusion} concludes the paper and gives future work.
